---
title: "Practical Machine Learning Course Project"
author: "Wesley Simons"
date: "July 16, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Wesley Simons Practical Machine Learning Course Project

###Loading the libraries and data. 
First we will load the required libraries, and the data files from the local machine. 
```{r libraries, results='hide', message=FALSE, warning=FALSE}
library(caret)
library(tidyverse)
library(rattle)

train.dat<-file.choose()
test.dat<-file.choose()
trainingSet<-read.csv(train.dat)
testingSet<-read.csv(test.dat)
```

###Preparing the data sets for analysis. 
Next we will create a data partition with the training set to build our model and test our prediction.
```{r partition, results='hide', message=FALSE}
set.seed(1234)
inTrain<- createDataPartition(trainingSet$classe, p=.75, list=FALSE)
wesTraining<-trainingSet[inTrain,]
wesTesting<-trainingSet[-inTrain,]
```

After our partitions are created we need to clean the data. 
```{r cleaning, results='hide'}
#remove the first two columns as they are not related to our analysis. 
wesTraining<-wesTraining[,-c(1:2)]
#remove all columns with more than 50% N/A values. 
na.names<-c(NULL)
for(i in 1:length(wesTraining)) {
  if(sum( is.na( wesTraining[, i] ) ) /nrow(wesTraining) >= .5) {
    na.names<-c(na.names,i)
  }
}
wesTraining<-wesTraining[,-na.names]
#remove the variables with near zero variance
nearzero<-nearZeroVar(wesTraining, saveMetrics = TRUE)
wesTraining<-wesTraining[,nearzero$nzv==FALSE]
#remove timestamp columns
wesTraining<-wesTraining[,-c(1:4)]
```
In the final data preparation step we will need to make sure that all of our data sets share the same variables. 
```{r match data sets}
#make training and testing sets the same
clean.names<-names(wesTraining)
wesTesting<-wesTesting[clean.names]
#make final testing set the same
testingSet<-testingSet[clean.names[-53]]
dim(wesTraining)
dim(wesTesting)
#the testingSet data.frame will have one less column as it does not contain the "classe" variable. 
dim(testingSet)
```

###Building the Models
####Decision Tree
We will evaluate a few different models to try and find one that accurately predicts the classe (form break) outcome. We will start with using the rpart package to build a decision tree. 
```{r rpart}
modfit.rpart<-train(classe~.,data=wesTrainin, method="rpart")

predict.rpart<-predict(modfit.rpart, newdata=wesTesting[,-57])
conf.rpart<-confusionMatrix(predict.rpart, wesTesting$classe)
conf.rpart$overall[1]
```
Unfortunately, our accuracy is very low, only ~47%.

####Random Forest
Since our decision tree accuracy was so poor, we will next try a random forest model. 
```{r rf}
modfit.rf<-train(classe~.,data=wesTraining, method="rf", trControl=trainControl(method="cv", number=4))

predict.rf<-predict(modfit.rf, newdata=wesTesting[,-57])
conf.rf<-confusionMatrix(predict.rf, wesTesting$classe)
conf.rf$overall[1]
```
The accuracy using the random forest model was much better, 99%! Let's look at the variables that our model found to be the most important, and how the accuracy changed as more predictors were included. Ultimately, our model only utilized 2 predictors, as this returned the highest accuracy.
```{r rf analysis}
varImp(modfit.rf)
plot(modfit.rf)
```
Given the high level of accuracy from our random forest model, we would predict that our out of sample error is 0.08%, or (1-.9991843)*100. 

###Predicting on the testing set
Finally, we will need to predict the classe variable for our testing set. 
```{r}
predict.test<-predict(modfit.rf, newdata = testingSet)
predict.test
```


